### 2023-01-28 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_interpreting_probabilities@77c553fff781e2cc1386dfe1493348f9f35114d5 🚀

Cataloging species interactions across space is a gargantuan task. At
the core of this challenge lies the spatiotemporal variability of
ecological networks (Poisot \emph{et al.} 2015), which makes documenting
the location and timing of interactions difficult. Indeed, it is not
sufficient to know if two species have the biological capacity to
interact to infer the realization of this interaction at a specific time
and space. Taking food webs as an example, a predator and its potential
prey must first co-occur on the same territory in order for a trophic
interaction to take place. They must then cross paths, which can happen
only if their relative abundances in the ecosystem is high enough and
their phenology matches. Finally, the interaction occurs only if the
predator has a desire to consume its prey and is able to capture and
ingest it. Environmental (e.g. temperature and presence of shelters) and
biological (e.g.~physiological state of both individuals and
availability of other prey species) factors contribute to this
variability by impacting species co-occurrence and the realization of
their interactions. In this context, it is unsurprising that
computational methods are being developed in ecology to help alleviate
the colossal sampling efforts required to document species interactions
across time and space (Strydom \emph{et al.} 2021).

The recognition of the variability of species interactions and the
emergence of numerical methods have led ecologists to rethink their
representation of ecological networks, slowly moving from a binary to a
probabilistic view of species interactions
(\textbf{Poisot2016StrProa?}). This has several benefits. For example,
probabilities represent the limit of our knowledge about species
interactions and can indicate the expected frequency of two species
interacting with each other. They are also very helpful in predictive
models when modeling the spatial distribution and temporal variability
of ecological networks, generating new ecological data, and identifying
priority sampling locations.

However, representing species interactions probabilistically can also be
challenging. Beyond methodological difficulties in estimating these
numbers, there are important conceptual challenges in defining what we
mean by ``probability of interactions.'' Because this ecological
representation is still in its infancy, there is no clear definition
found in the literature. In this contribution, we outline different ways
to define and interpret interactions probabilities in network ecology
and propose an approach to thinking about them. These definitions mostly
depend on the study system (e.g.~local network or metaweb) and on the
method used to generate them. We show that different definitions can
have different ecological implications, especially related to spatial,
temporal, and taxonomic scaling. Although we will focus on food webs,
our observations and advice can be applied to all types of ecological
networks, from plant-pollinator to host-parasite networks. Overall, we
argue that probabilities should be better documented, defined
mathematically, and used with caution when describing species
interactions.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-McLeod2020EffSpe}{}%
McLeod, A.M., Leroux, S.J. \& Chu, C. (2020). Effects of species traits,
motif profiles, and environment on spatial variation in multi-trophic
antagonistic networks. \emph{Ecosphere}, 11, e03018.

@@

>>>>

Cataloging species interactions across space is a gargantuan task. At
the core of this challenge lies the spatiotemporal variability of
ecological networks (Poisot \emph{et al.} 2012, 2015), which makes
documenting the location and timing of interactions difficult. Indeed,
it is not sufficient to know if two species have the biological capacity
to interact to infer the realization of their interaction at a specific
time and space {[}TK{]}. Taking food webs as an example, a predator and
its potential prey must first co-occur on the same territory in order
for a trophic interaction to take place (Blanchet \emph{et al.} 2020).
They must then encounter, which is conditional on their relative
abundances in the ecosystem and the matching of their phenology (Poisot
\emph{et al.} 2015). Finally, the interaction occurs only if the
predators have a desire to consume their prey and are able to capture
and ingest them (Pulliam 1974). Environmental (e.g.~temperature and
presence of shelters) and biological (e.g.~physiological state of both
species and availability of other prey species) factors contribute to
this variability by impacting species co-occurrence {[}TK{]} and the
realization of their interactions {[}TK{]}. In this context, it is
unsurprising that computational methods are being developed in ecology
to help alleviate the colossal sampling efforts required to document
species interactions across time and space (Strydom \emph{et al.} 2021).
Having a better portrait of species interactions and the emerging
structure of food webs is important since it lays the groundwork for
understanding the functioning {[}TK{]}, dynamics {[}TK{]}, and
resilience {[}TK{]} of ecosystems worldwide.

The recognition of the variability of species interactions and the
emergence of numerical methods have led ecologists to rethink their
representation of ecological networks, slowly moving from a binary to a
probabilistic view of species interactions
(\textbf{Poisot2016StrProa?}). This has several benefits. For example,
probabilities represent the limit of our knowledge about species
interactions and can indicate the expected frequency of two species
interacting with each other {[}TK{]}. They are also very helpful in
predictive models when modeling the spatial distribution of species
(Cazelles \emph{et al.} 2016) and the temporal variability of ecological
networks {[}TK{]}, generating new ecological data (e.g., Strydom
\emph{et al.} 2022), and identifying priority sampling locations (see
Andrade-Pacheco \emph{et al.} 2020 for an ecological example of a
sampling optimization problem). Moreover, the high rate of false
negatives in ecological network data, resulting from the difficulty of
witnessing interactions between rare species, makes it hard to interpret
non-observations of species interactions ecologically (Catchen \emph{et
al.} 2023). Using probabilities instead of yes-no interactions accounts
for these observation errors; in that case, only forbidden interactions
(Jordano \emph{et al.} 2003; Olesen \emph{et al.} 2010) would have a
probability value of zero (but see Gonzalez-Varo \& Traveset 2016).

However, representing species interactions probabilistically can also be
challenging. Beyond methodological difficulties in estimating these
numbers, there are important conceptual challenges in defining what we
mean by ``probability of interactions.'' To the best of our knowledge,
because the building blocks of this mathematical representation of food
webs are still being laid, there is no clear definition found in the
literature. This is worrisome, since working with probabilistic species
interactions without clear guidelines could be misleading as much for
field ecologists as for computational ecologists who use and generate
these data. In this contribution, we outline different ways to define
and interpret interactions probabilities in network ecology and propose
an approach to thinking about them. These definitions mostly depend on
the study system (e.g.~local network or metaweb) and on the method used
to generate them. We show that different definitions can have different
ecological implications, especially regarding spatial, temporal, and
taxonomic scaling. Although we will focus on food webs, our observations
and advice can be applied to all types of ecological networks, from
plant-pollinator to host-parasite networks. Specifically, we argue that
probabilities should be better documented, defined mathematically, and
used with caution when describing species interactions.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Andrade-Pacheco2020FinHot}{}%
Andrade-Pacheco, R., Rerolle, F., Lemoine, J., Hernandez, L., Meïté, A.,
Juziwelo, L., \emph{et al.} (2020). Finding hotspots: Development of an
adaptive spatial sampling approach. \emph{Scientific Reports}, 10,
10939.

\leavevmode\hypertarget{ref-Blanchet2020CooNot}{}%
Blanchet, F.G., Cazelles, K. \& Gravel, D. (2020). Co-occurrence is not
evidence of ecological interactions. \emph{Ecology Letters}, 23,
1050--1063.

\leavevmode\hypertarget{ref-Catchen2023MisLin}{}%
Catchen, M.D., Poisot, T., Pollock, L.J. \& Gonzalez, A. (2023). The
missing link: Discerning true from false negatives when sampling species
interaction networks.

\leavevmode\hypertarget{ref-Cazelles2016TheSpe}{}%
Cazelles, K., Araujo, M.B., Mouquet, N. \& Gravel, D. (2016). A theory
for species co-occurrence in interaction networks. \emph{Theoretical
Ecology}, 9, 39--48.

\leavevmode\hypertarget{ref-Gonzalez-Varo2016LabLim}{}%
Gonzalez-Varo, J.P. \& Traveset, A. (2016). The Labile Limits of
Forbidden Interactions. \emph{Trends in Ecology \& Evolution}, 31,
700--710.

\leavevmode\hypertarget{ref-Jordano2003InvPro}{}%
Jordano, P., Bascompte, J. \& Olesen, J.M. (2003). Invariant properties
in coevolutionary networks of plantanimal interactions. \emph{Ecology
Letters}, 6, 69--81.

\leavevmode\hypertarget{ref-McLeod2020EffSpe}{}%
McLeod, A.M., Leroux, S.J. \& Chu, C. (2020). Effects of species traits,
motif profiles, and environment on spatial variation in multi-trophic
antagonistic networks. \emph{Ecosphere}, 11, e03018.

\leavevmode\hypertarget{ref-Olesen2010MisFor}{}%
Olesen, J.M., Bascompte, J., Dupont, Y.L., Elberling, H., Rasmussen, C.
\& Jordano, P. (2010). Missing and forbidden links in mutualistic
networks. \emph{Proceedings of the Royal Society B: Biological
Sciences}, 278, 725--732.

\leavevmode\hypertarget{ref-Poisot2012DisSpe}{}%
Poisot, T., Canard, E., Mouillot, D., Mouquet, N. \& Gravel, D. (2012).
The dissimilarity of species interaction networks. \emph{Ecology
Letters}, 15, 1353--1361.

@@

\leavevmode\hypertarget{ref-Pulliam1974TheOpt}{}%
Pulliam, H.R. (1974). On the Theory of Optimal Diets. \emph{The American
Naturalist}, 108, 59--74.

\leavevmode\hypertarget{ref-Strydom2022Food}{}%
Strydom, T., Bouskila, S., Banville, F., Barros, C., Caron, D., Farrell,
M.J., \emph{et al.} (2022). Food web reconstruction through phylogenetic
transfer of low-rank network representation. \emph{Methods in Ecology
and Evolution}, 13.
